%% AASTeX v7.0.1 (released May 29, 2025)
%% Classfile: aastex701.cls
%% Download from: https://journals.aas.org/aastex-package-for-manuscript-preparation/
%% Or use via Overleaf (pre-installed)
\documentclass[preprint]{aastex701-1/aastex701}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\begin{document}

\title{Prism: A Physics-Constrained Neural Emulator for Stellar Spectra}

\author[0009-0006-8079-688X]{Aneesh Shastri}
\affiliation{Department of Computational Engineering, Cepheid,
Indian Institute of Technology Hyderabad,
Sangareddy, Telangana 502284, India}
\email{co25btech11004@iith.ac.in}

\begin{abstract}
We present a physics-constrained generative neural network emulator for stellar spectra
trained directly on real, high signal-to-noise observations from the APOGEE DR17 survey.
Unlike prior approaches that rely on synthetic spectral grids, our model is trained on
149,966 stars with signal-to-noise ratio $>100$ using stellar parameters and chemical
abundances from the ASPCAP pipeline as labels, spanning a wavelength range of 1514--1694\,nm.
The model predicts optical depth through four physically motivated branches---a continuum
branch, a CNO molecular branch, a thermodynamic state branch, and a chemical expert branch
with element-specific weighted spectral masks sourced from ASPCAP's \texttt{.filt}
files---and converts to flux via the Beer-Lambert law. Physics-based disentanglement is
enforced both architecturally, by restricting each element's spectral sensitivity to its
permitted spectral windows, and during training, via a stochastic Jacobian penalty that
penalizes spectral response at forbidden pixels. We evaluate model performance using a
reduced $\chi^2$ statistic with full error propagation and compare against four baselines:
The Payne, an Expanded Payne, a baseline CNN, and a regularized Payne (RegPayne). PRISM
achieves a median reduced $\chi^2_\mathrm{red} = 1.328$ on a held-out test set of
$\sim$9,500 stars, outperforming all baselines. An information-leakage test confirms that
PRISM's gradient response at forbidden pixels is substantially suppressed relative to
unconstrained baselines. Principal component analysis confirms that the generated spectra
faithfully reproduce the manifold of real APOGEE observations. Sensitivity gradients
$\partial F / \partial [\mathrm{Al/H}]$ are identically zero at all non-permitted
wavelengths, demonstrating physically meaningful disentanglement.
\end{abstract}

\keywords{stars: abundances --- stars: fundamental parameters --- methods: data analysis ---
techniques: spectroscopic --- surveys}

%------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
%------------------------------------------------------------

The chemical fingerprint encoded in a stellar photosphere reflects the conditions of the
interstellar medium at the time and place of that star's formation, as well as the integrated
nucleosynthetic enrichment from preceding generations of supernovae and asymptotic giant
branch stars. Because photospheric abundances change negligibly over a stellar lifetime,
high-fidelity abundance measurements from large spectroscopic surveys constitute a powerful
fossil record of Galactic chemical evolution \citep{Freeman2002,BlandHawthorn2016}.

This goal has only recently become tractable as survey facilities deliver high-resolution
($R \gtrsim 20{,}000$) spectra for $\sim 10^5$--$10^6$ stars across all Galactic components
\citep{Majewski2017}. APOGEE (Apache Point Observatory Galactic Evolution Experiment)
in particular has assembled one of the largest homogeneous near-infrared spectroscopic
datasets in existence, with Data Release 17 containing labels derived for hundreds of
thousands of giant stars \citep{Abdurrouf2022}.

Automated pipelines such as ASPCAP (APOGEE Stellar Parameter and Chemical Abundance
Pipeline) have made survey-scale stellar characterization possible, yet physics-based model
spectra are known to be incomplete: one-dimensional model atmospheres, the local
thermodynamic equilibrium approximation, and imperfect atomic and molecular line lists
all introduce systematic residuals \citep{GarciaperezASPCAP2016}. These limitations become
especially acute at low signal-to-noise ratios, where classical $\chi^2$ fitting degrades
rapidly.

Data-driven models offer a complementary approach. The Cannon \citep{Ness2015Cannon}
demonstrated that a quadratic generative model trained on high-fidelity reference spectra
can transfer stellar labels to lower signal-to-noise data with competitive precision. The
Cannon 2 \citep{Casey2016Cannon2} extended this to 17 labels simultaneously by introducing
$\ell_1$ regularization that enforces sparsity in the model's spectral coefficients, yielding
physically interpretable spectral derivatives. The Payne \citep{Ting2019Payne} replaced the
polynomial basis with a multi-layer perceptron trained on grids of synthetic spectra,
achieving high accuracy on APOGEE observations. More recently, TransformerPayne
\citep{TransformerPayne2025} employed large transformer architectures over the same
synthetic-grid paradigm, reporting mean absolute errors of $\sim 10^{-3}$ on held-out
synthetic data.

A common thread in nearly all prior emulators is reliance on synthetic spectral grids
for training. While such grids are noise-free by construction, they encode all the
approximations of the underlying physics model and, critically, are evaluated on real data
that differ systematically from their training distribution. Furthermore, most neural emulators
operate as black boxes: the learned mapping from stellar parameters to predicted flux offers
little interpretable insight into which spectral features are being used, leaving open the
question of parameter cross-talk (entanglement).

Here we present a generative neural emulator trained entirely on real APOGEE DR17
spectra. Our model is distinguished by three design principles. First, it predicts optical
depth rather than flux directly, using the Beer-Lambert law to recover flux from summed
contributions of physically separated branches. Second, element-specific spectral masks
enforce strict disentanglement: each chemical abundance label is permitted to influence only
those pixels lying within 0.2\,nm of its known atomic lines, as compiled in
\citet{Shetrone2015} and \citet{Smith2021DR17}. Third, because the training set is composed
of real observations with realistic noise, continuum normalization artifacts, and sensor gaps,
the emulator inherits the noise properties of APOGEE data without requiring a synthetic
bridge.

Throughout this work we use the term \emph{label} for any element of the input
parameter vector, including stellar parameters ($T_\mathrm{eff}$, $\log g$, etc.) and
chemical abundances ([$X$/Fe] or [$X$/H]).

%------------------------------------------------------------
\section{Data}
\label{sec:data}
%------------------------------------------------------------

\subsection{APOGEE DR17}
\label{subsec:data_apogee}

We use spectra and labels from APOGEE Data Release 17 \citep{Abdurrouf2022}. The APOGEE
spectrograph covers the $H$-band (1514--1694\,nm) at resolution $R \approx 22{,}500$.
Continuum normalization and spectral stacking are performed by the ASPCAP pipeline before
data release \citep{GarciaperezASPCAP2016}. Sensor gaps (pixels with zero flux) are masked
throughout all processing steps.

Our working sample comprises 149,966 stars selected to have signal-to-noise ratio
SNR $> 100$ in the stacked, pseudo-continuum-normalized spectra. Stellar parameters and
23 chemical abundance labels are taken from ASPCAP (see Table~\ref{tab:labels}). Six core
stellar parameters are used: effective temperature $T_\mathrm{eff}$, surface gravity $\log g$,
overall metallicity $[M/\mathrm{H}]$, microturbulent velocity $v_\mathrm{micro}$,
macroturbulent velocity $v_\mathrm{macro}$, and projected rotational velocity $v \sin i$.
The 17 elemental abundances span the CNO group, the $\alpha$-capture, odd-Z, and Fe-peak
families.

\begin{deluxetable}{ll}
\tablecaption{Labels used in model training \label{tab:labels}}
\tablewidth{0pt}
\tablehead{\colhead{Group} & \colhead{Labels}}
\startdata
Core parameters & $T_\mathrm{eff}$, $\log g$, $[M/\mathrm{H}]$, $v_\mathrm{micro}$,
                  $v_\mathrm{macro}$, $v\sin i$ \\
CNO group        & $[\mathrm{C/Fe}]$, $[\mathrm{N/Fe}]$, $[\mathrm{O/Fe}]$ \\
Iron peak        & $[\mathrm{Fe/H}]$, $[\mathrm{Mn/Fe}]$, $[\mathrm{Ni/Fe}]$,
                  $[\mathrm{Cr/Fe}]$, $[\mathrm{V/Fe}]$, $[\mathrm{Co/Fe}]$ \\
$\alpha$-capture & $[\mathrm{Mg/Fe}]$, $[\mathrm{Si/Fe}]$, $[\mathrm{Ca/Fe}]$,
                  $[\mathrm{Ti/Fe}]$, $[\mathrm{S/Fe}]$ \\
Odd-Z            & $[\mathrm{Al/Fe}]$, $[\mathrm{K/Fe}]$, $[\mathrm{Na/Fe}]$
\enddata
\end{deluxetable}

\subsection{Preprocessing and Label Imputation}
\label{subsec:preproc}

Missing label values are common in APOGEE because the ASPCAP pipeline flags individual
abundance fits when signal is insufficient. Rather than discard affected stars, we recover
missing values via iterative imputation using a Bayesian Ridge regressor that learns
inter-label correlations (e.g., the tight correlation between $[\mathrm{Fe/H}]$ and
$[M/\mathrm{H}]$) and predicts missing entries from available ones. After imputation,
all [$X$/Fe] ratios are converted to [$X$/H] by adding the imputed $[\mathrm{Fe/H}]$
value.

Four engineered features supplement the raw labels to improve the model's ability to
decode stellar physics:
\begin{enumerate}
\item $5040/T_\mathrm{eff}$ -- encodes the Boltzmann exponential dependence of line opacity
      on excitation temperature;
\item $[\mathrm{C/H}] - [\mathrm{O/H}]$ -- the C/O ratio, which governs the balance
      between CO molecule formation and free atomic C and O;
\item $v_\mathrm{broad} = \sqrt{v_\mathrm{macro}^2 + (v\sin i)^2}$ -- a single
      combined broadening term;
\item $\log P_e = 0.5[\mathrm{M/H}]+0.5\log{g}$ -- a first-order approximation of electron pressure derived from
      metallicity and surface gravity, which modulates continuous opacity.
\end{enumerate}

All labels (raw and engineered) are $z$-score normalized using training-set statistics.
Spectral fluxes are used as delivered by ASPCAP (already pseudo-continuum normalized).

%------------------------------------------------------------
\section{Model Architecture}
\label{sec:model}
%------------------------------------------------------------

\subsection{Optical-depth Formulation}
\label{subsec:tau}

Rather than regressing directly to continuum-normalized flux, our model predicts the
effective optical depth $\tau(\lambda)$ and converts to flux via the Beer-Lambert law:
\begin{equation}
F(\lambda) = k \, e^{-\tau(\lambda)},
\label{eq:beer}
\end{equation}

where $k = \mathrm{k(\lambda)}$ is a scaling factor, which is the output of the continuum branch.We chose a scaling factor instead of another optical depth function to combat the issues with the ASPCAP continuum normalization (which can produce normalized fluxes slightly above unity near absorption features).

The total optical depth is additive in contributions from
physically distinct components:
\begin{equation}
\tau(\lambda) =  \tau_\mathrm{CNO}(\lambda)
               + \sum_X \tau_X(\lambda),
\label{eq:tau_sum}
\end{equation}
where the sum runs over all atomic species $X$ in Table~\ref{tab:labels} (excluding the
CNO group), and each term is produced by a dedicated network branch.

\subsection{Network Branches}
\label{subsec:branches}

\subsubsection{Continuum Branch}
The continuum branch receives the global parameters:
$T_{\text{eff}},[\mathrm{M/H}],\log{g},v_{\text{macro}},v_{\text{micro}},v\sin{i}$.
(Along with these, it also receives all our engineered features)

Using these, it generates a coarse 268-pixel continuum that is progressively upsampled via alternating
upsampling and Conv1D layers to the full 8575-pixel grid. The final convolutional layer
uses a kernel of width 51 pixels, ensuring the branch produces smoothly varying pseudo-
continuum and is structurally incapable of fitting sharp spectral lines.

\subsubsection{CNO (Molecular) Branch}
The CNO branch receives $[\mathrm{C/H}]$, $[\mathrm{N/H}]$, $[\mathrm{O/H}]$,
$T_\mathrm{eff}$, and $\log g$. Because molecular line lists (CO, CN, OH, etc.) are
incomplete and thousands of transitions contribute at every wavelength, this branch is
intentionally unconstrained by pixel masks. It uses stacked dense layers to predict
$\tau_\mathrm{CNO}$ across the full wavelength array.

\subsubsection{Thermodynamic State Branch}
The thermodynamic state branch distills all global and CNO parameters into a compact
latent state vector of dimension 8. This state vector is passed to the chemical expert
branch rather than the raw label values, preventing the chemical branch from using
$[M/\mathrm{H}]$ as a proxy for individual elemental abundances. This architectural
choice is the primary mechanism by which the model enforces physically correct
label--flux correlations.

\subsubsection{Chemical Expert Branch}
Each atomic species $X$ has its own dedicated expert block. The block concatenates the
thermodynamic state vector with a single scalar $[X/\mathrm{H}]$ and passes the result
through two dense layers (64 units each), followed by a sparse spectral window projection.
The spectral windows are defined by ASPCAP's \texttt{.filt} files
\citep{GarciaperezASPCAP2016,Abdurrouf2022},
which provide continuous-valued significance weights for each pixel--species pair rather
than binary on/off masks. The \texttt{SparseProjector} layer scatters the expert's local
optical depth prediction into the full 8575-pixel array, weighting each pixel by its
species-specific significance. Pixels with zero weight are structurally excluded from the
output:
\begin{equation}
\tau_X(\lambda_i) = w_{X,i} \, f_X\bigl(\mathbf{s}, [X/\mathrm{H}]\bigr),
\label{eq:mask}
\end{equation}
where $w_{X,i}$ is the ASPCAP window weight for species $X$ at pixel $i$,
$\mathbf{s}$ is the thermodynamic state vector, and $f_X$ is the expert sub-network.

This weighting is a hard mathematical constraint: gradients cannot propagate through
zero-weight pixels, so the network is provably unable to attribute spectral signal at a
non-permitted wavelength to species $X$.

\subsection{Training}
\label{subsec:training}

The model is trained end-to-end using the Adam optimizer with gradient clipping
($\|\nabla\|_{\max} = 1.0$) and an initial learning rate of $10^{-3}$. The APOGEE
dataset is split into training (120{,}000 stars), validation (20{,}000 stars), and test
($\sim$9{,}500 stars) sets. Sensor-gap pixels (identified by zero flux or zero inverse
variance) are masked from the loss throughout. Training uses a batch size of 256 and
runs for up to 150 epochs. Three callbacks regulate convergence:
\texttt{ReduceLROnPlateau} (patience 2, factor 0.5), \texttt{EarlyStopping} (patience
10, monitoring validation loss), and \texttt{ModelCheckpoint} saving the best-performing
weights.

\subsubsection{Stochastic Jacobian Penalty}
\label{subsec:jacobian_penalty}

In addition to the spectral-mask constraint, we enforce disentanglement during training
via a stochastic Jacobian penalty. A \texttt{StochasticPerturbation} layer randomly
selects one constrained abundance label per batch, adds a small perturbation
($\epsilon = 0.05$) to that label, and runs a ``shadow pass'' through the model. A
\texttt{PhysicsLossLayer} then computes the difference between the clean and perturbed
outputs at wavelengths where the perturbed element has zero mask weight (forbidden
pixels). This difference is penalized with a coefficient $\lambda_J = 2500$, encouraging
the model to produce zero spectral response at forbidden wavelengths even when training
data might otherwise permit leakage through shared correlations.

%------------------------------------------------------------
\section{Loss and Evaluation Metrics}
\label{sec:loss}
%------------------------------------------------------------

\subsection{Training Loss: Spectral Pixel Loss (\texttt{spl\_loss})}
\label{subsec:spl_loss}

The training objective is a weighted mean squared error computed per pixel:
\begin{equation}
\mathcal{L}_\mathrm{spl} = \frac{1}{N_\mathrm{valid}} \sum_{i}
  w_i \bigl(F_{\mathrm{pred},i} - F_{\mathrm{obs},i}\bigr)^2,
\label{eq:spl_loss}
\end{equation}
where the per-pixel weight is
\begin{equation}
w_i = \frac{10^{-5}}{1/\mathit{ivar}_i + 10^{-5}},
\label{eq:spl_weight}
\end{equation}
and $\mathit{ivar}_i$ is the APOGEE inverse variance at pixel $i$. The weight smoothly
transitions from zero (where $\mathit{ivar} = 0$, i.e.\ sensor gaps) to unity (where
signal-to-noise is high), and the term $10^{-5}$ acts as a variance floor that prevents
numerical instabilities. Pixels with observed flux below a bad-pixel cutoff
($F_\mathrm{obs} < 10^{-3}$) are additionally zeroed to exclude defective pixels.

\subsection{Monitoring Metric: Reduced $\chi^2$}
\label{subsec:chi2_metric}

During training a Keras metric computes a batch-level estimate of the reduced
$\chi^2$:
\begin{equation}
\hat{\chi}^2_\mathrm{red} = \frac{1}{\nu} \sum_{i}
  \mathit{ivar}_i \, \bigl(F_{\mathrm{pred},i} - F_{\mathrm{obs},i}\bigr)^2,
\label{eq:chi2_batch}
\end{equation}
where $\nu$ is the number of valid pixels per star and we use raw
$\mathit{ivar}$ (not the scaled weight of Equation~\ref{eq:spl_weight}). This metric
provides a physically interpretable goodness-of-fit during training without affecting
gradient updates.

\subsection{Test-time Evaluation: Full Reduced $\chi^2$}
\label{subsec:chi2_full}

For final model evaluation on the held-out test set, we compute a more rigorous per-star
reduced $\chi^2$ with full error propagation:
\begin{equation}
\chi^2_{\mathrm{red},\star} = \frac{1}{\nu_\star} \sum_{i=1}^{N_\mathrm{pix}}
  \frac{\bigl(F_{\mathrm{pred},i} - F_{\mathrm{obs},i}\bigr)^2}
       {\sigma^2_{\mathrm{instr},i} + \sigma^2_{\mathrm{model},i}
        + \sigma^2_\mathrm{floor}},
\label{eq:chi2_full}
\end{equation}
where $\sigma^2_\mathrm{instr} = 1/\mathit{ivar}$ is the instrumental variance,
$\sigma^2_\mathrm{floor} = 10^{-5}$ prevents division by zero, and
$\sigma^2_\mathrm{model}$ captures uncertainty propagated through the model via
finite-difference gradients ($\epsilon = 10^{-3}$) applied to each label in turn and
weighted by the ASPCAP label uncertainties. The degrees of freedom $\nu_\star$ equal
the number of valid pixels minus the number of input labels.

%------------------------------------------------------------
\section{Validation}
\label{sec:validation}
%------------------------------------------------------------

We evaluate the model on a held-out test set of $\sim$9{,}500 stars
(indices 140{,}000--149{,}500) against four baselines trained on the same data split:
The Payne, an Expanded Payne (larger hidden layers), a baseline CNN
(convolutional architecture, no physics constraints), and RegPayne (a regularized
variant trained with a two-phase Jacobian penalty schedule).

\subsection{Reduced $\chi^2$ Analysis}
\label{subsec:chi2_results}

Table~\ref{tab:chi2} reports the test-set reduced $\chi^2$ for each model computed
via Equation~(\ref{eq:chi2_full}). PRISM achieves the lowest median
$\chi^2_\mathrm{red}$, followed closely by RegPayne and Expanded Payne. The unconstrained
CNN performs worst, with both high median and a heavy right tail (95th percentile
$> 11$).

\begin{deluxetable}{lrrr}
\tablecaption{Reduced $\chi^2$ on the held-out test set \label{tab:chi2}}
\tablewidth{0pt}
\tablehead{
  \colhead{Model} &
  \colhead{Median $\chi^2_\mathrm{red}$} &
  \colhead{Mean} &
  \colhead{95th \%}
}
\startdata
PRISM          & 1.328 & 1.566 &  2.902 \\
RegPayne       & 1.347 & 1.621 &  3.022 \\
Expanded Payne & 1.412 & 1.738 &  3.367 \\
The Payne      & 1.741 & 2.274 &  4.771 \\
CNN            & 2.923 & 4.336 & 11.279
\enddata
\tablecomments{All models are evaluated on the same $\sim$9{,}500-star test set with
identical error-propagation methodology.}
\end{deluxetable}

We further examine $\chi^2_\mathrm{red}$ as a function of $T_\mathrm{eff}$,
$\log g$, $[M/\mathrm{H}]$, and SNR by plotting the binned median and standard
deviation for all five models (Figure~\ref{fig:chi2_binned}). PRISM maintains
$\chi^2_\mathrm{red}$ closest to unity across all parameter ranges.

\subsection{$\chi^2$ by Stellar Type}
\label{subsec:chi2_stellar}

To probe performance across astrophysically distinct populations, we partition the
test set into three categories and compute per-group statistics
(Table~\ref{tab:chi2_stellar}):

\begin{deluxetable*}{lrrrrrr}
\tablecaption{Median reduced $\chi^2$ by stellar type \label{tab:chi2_stellar}}
\tablewidth{0pt}
\tablehead{
  \colhead{Stellar Group} &
  \colhead{$N$} &
  \colhead{PRISM} &
  \colhead{Payne} &
  \colhead{Exp.\ Payne} &
  \colhead{CNN} &
  \colhead{RegPayne}
}
\startdata
Cool Giants ($T_\mathrm{eff}<4500$, $\log g<3.5$) & 3045 & 1.561 & 1.996 & 1.614 & 4.311 & 1.550 \\
Sun-like ($5000<T_\mathrm{eff}<6000$, $\log g>4$)  &  217 & 1.435 & 1.815 & 1.610 & 3.366 & 1.461 \\
Hot Stars ($T_\mathrm{eff}>6500$)                   &   66 & 1.828 & 2.374 & 2.017 & 7.053 & 1.662
\enddata
\tablecomments{RegPayne slightly outperforms PRISM in the cool-giant and hot-star bins;
PRISM leads in the sun-like category. Both physics-constrained models significantly
outperform the unconstrained CNN.}
\end{deluxetable*}

\subsection{Information Leakage Test}
\label{subsec:leakage}

An information-leakage test quantifies how much gradient power an element produces at
wavelengths where it has no known spectral lines. For each of the 14 atomic species, we
compute the finite-difference gradient $\partial F / \partial [X/\mathrm{H}]$ over the
full test set and partition pixels into \emph{allowed} (non-zero ASPCAP window weight)
and \emph{forbidden} (zero weight) regions using the \texttt{apogee\_mask.npy}. The
leakage ratio is defined as $\bar{|\nabla|}_\mathrm{forbidden} \,/\,
\bar{|\nabla|}_\mathrm{allowed}$; a value well below unity indicates that the model
concentrates its sensitivity on the correct spectral windows.

PRISM achieves leakage ratios of $\sim$0.36--0.94 across elements
(Figure~\ref{fig:leakage}), substantially below 1.0 for most species, confirming that
the architectural mask and Jacobian penalty together suppress spectral response at
forbidden pixels. All four unconstrained baselines exhibit leakage ratios
$\sim$0.94--1.05, consistent with uniform gradient power across all pixels and an
absence of disentanglement.

\subsection{Manifold Coverage: PCA Analysis}
\label{subsec:pca}

To assess whether the emulator generates spectra whose statistical properties match those
of real APOGEE observations, we performed a principal component analysis (PCA) on a
matched sample of real and generated spectra (with sensor-gap artifacts masked for a fair
comparison). The first two principal components explain 54.2\% and 18.0\% of the variance,
respectively. As shown in Figure~\ref{fig:pca}, the generated spectra occupy the same
two-dimensional projection as the real APOGEE spectra, demonstrating that the model
faithfully reproduces the data manifold. The cumulative explained variance for the
generated spectra rises slightly faster than for the real data, consistent with the fact
that our deterministic model does not reproduce observational noise (which distributes
variance broadly across all components).

\subsection{Physics Disentanglement: Gradient Test}
\label{subsec:gradient}

A central claim of our model is that chemical abundance labels affect the predicted flux
only at physically permitted wavelengths. We verify this directly by computing the
sensitivity gradient $\partial F / \partial [\mathrm{Al/H}]$ at every pixel for a
representative test star. Figure~\ref{fig:gradient} shows that this gradient is
indistinguishable from zero at all wavelengths except those corresponding to known
Al absorption lines, with two prominent non-zero peaks visible near the permitted Al
transitions at $\sim$1675\,nm. No leakage to Fe, CO, or other features is detected.

This behavior is a direct consequence of the weighted spectral mask in
Equation~(\ref{eq:mask}): because the forward computation for element $X$ at zero-weight
pixels is identically zero, the backpropagated gradient is identically zero at those
pixels by the chain rule. The gradient test therefore confirms that the architectural
disentanglement is operating as intended.

%------------------------------------------------------------
\section{Comparison with Prior Work}
\label{sec:comparison}
%------------------------------------------------------------

\subsection{The Cannon and The Cannon 2}
\label{subsec:cannon}

The Cannon \citep{Ness2015Cannon} and The Cannon 2 \citep{Casey2016Cannon2} are
data-driven models that fit a polynomial (quadratic) generative function relating stellar
labels to spectral fluxes. The Cannon 2 uses $\ell_1$ regularization to enforce sparsity
in the spectral derivatives, which---like our spectral masks---prevents label coefficients
from acquiring unphysical non-zero values at all wavelengths. Our model shares this
philosophy of enforcing physical sparsity but improves upon it in two respects: first,
the sparsity in our model is mathematically exact (hard masking) rather than
approximate (soft regularization); second, our non-linear neural architecture can capture
higher-order interactions between labels and spectra that a quadratic model cannot represent.

\subsection{The Payne}
\label{subsec:payne}

The Payne \citep{Ting2019Payne} is a multi-layer perceptron emulator trained on 2000
synthetic spectrograms with 7500 pixels, achieving a root-mean-square error of $10^{-3}$
on synthetic validation data. Because it is trained exclusively on synthetic spectra, it
is an \emph{ab initio} model: it can, in principle, return labels on any physical scale,
but it encodes all the approximations of the theoretical atmosphere grid. By contrast,
our model is trained on real APOGEE spectra and therefore learns the empirical noise floor
and calibration characteristics of the instrument.

We trained a Payne-architecture baseline on the same APOGEE data and data split as PRISM.
On our held-out test set, the Payne baseline achieves a median $\chi^2_\mathrm{red}
= 1.741$, compared with PRISM's $1.328$---a 24\% improvement. The Payne baseline also
suffers from leakage ratios near unity for all atomic species, confirming that it does not
disentangle abundance labels from each other. Our architecture explicitly prevents such
cross-talk through the weighted spectral masks and Jacobian penalty.

\subsection{TransformerPayne}
\label{subsec:transformerpayne}

TransformerPayne \citep{TransformerPayne2025} employs large transformer attention
mechanisms to capture long-range correlations across the full spectrum, again trained on
synthetic grids, and reports an MAE of $\sim 10^{-3}$ on a held-out synthetic dataset.
While this represents excellent performance on synthetic data, it requires substantially
more computational resources than our model and does not include real-data validation.
Our approach achieves comparable fitting accuracy on real observations with lower
computational cost and with physically enforced interpretability.

\subsection{Baseline Comparison: Expanded Payne, CNN, and RegPayne}
\label{subsec:baselines}

In addition to the original Payne architecture, we trained three further baselines on
the same data split:
\begin{enumerate}
\item \textbf{Expanded Payne} --- a wider MLP with larger hidden layers,
      increasing parameter capacity without adding physics constraints;
\item \textbf{CNN} --- a convolutional architecture that predicts spectra directly from
      labels, with no spectral masks or optical-depth decomposition;
\item \textbf{RegPayne} --- a Payne-like model trained with a two-phase Jacobian penalty
      schedule (phase 1: 5 epochs, $\lambda_J = 0$; phase 2: full training,
      $\lambda_J = 400$) using the \texttt{build\_disentangled\_model} framework.
\end{enumerate}
Results are summarized in Tables~\ref{tab:chi2} and \ref{tab:chi2_stellar}. RegPayne
achieves $\chi^2_\mathrm{red}$ performance close to PRISM, suggesting that even
moderate Jacobian regularization improves spectral fit quality. However, in the
information-leakage test (Section~\ref{subsec:leakage}), RegPayne's per-element leakage
ratios remain near unity, indicating that its weaker penalty is insufficient to enforce
true disentanglement.

\subsection{Summary of Comparison}

Table~\ref{tab:comparison} summarizes the key distinctions between our model and
representative prior work.

\begin{deluxetable*}{lcccr}
\tablecaption{Comparison of stellar spectral emulators \label{tab:comparison}}
\tablewidth{0pt}
\tablehead{
  \colhead{Model} &
  \colhead{Training data} &
  \colhead{Disentanglement} &
  \colhead{Interpretable} &
  \colhead{Median $\chi^2_\mathrm{red}$}
}
\startdata
The Cannon 2     & Real (APOGEE DR12) & Soft ($\ell_1$)   & Yes (derivatives) & ---   \\
The Payne        & Synthetic grid     & None              & No (black box)    & ---   \\
TransformerPayne & Synthetic grid     & None              & No (black box)    & ---   \\
\hline
Payne (ours)     & Real (APOGEE DR17) & None              & No                & 1.741 \\
Expanded Payne   & Real (APOGEE DR17) & None              & No                & 1.412 \\
CNN              & Real (APOGEE DR17) & None              & No                & 2.923 \\
RegPayne         & Real (APOGEE DR17) & Soft (Jacobian)   & Partial           & 1.347 \\
\textbf{PRISM}   & Real (APOGEE DR17) & Hard (mask + pen.) & Yes (gradients)  & \textbf{1.328}
\enddata
\tablecomments{Top section: published models compared qualitatively. Bottom section:
our baselines trained on the same data, with median $\chi^2_\mathrm{red}$ from
Table~\ref{tab:chi2}. Dashes indicate values not reported in the original publications
for real APOGEE data.}
\end{deluxetable*}

%------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}
%------------------------------------------------------------

We have presented a neural emulator for stellar spectra that combines the advantages of
data-driven modeling---training on real observations, robustness to noise and calibration
artifacts---with hard physical constraints that guarantee interpretable, disentangled label
sensitivities. The architectural separation of the spectrum into continuum, molecular, and
atomic contributions mirrors the physical origin of these spectral components and
provides a natural pathway for model interpretation.

The gradient test presented in Section~\ref{subsec:gradient} and the information-leakage
test in Section~\ref{subsec:leakage} constitute direct, pixel-level validations of
disentanglement that are not available from standard metrics such as cross-validation loss
or abundance precision. We argue that such tests should become standard practice for any
generative spectral model that claims interpretability. The stochastic Jacobian penalty
introduced in Section~\ref{subsec:jacobian_penalty} provides a training-time mechanism to
reinforce architectural disentanglement, and our comparison with RegPayne suggests that
stronger penalties combined with hard masks yield materially better leakage suppression
than weaker penalties alone.

Several limitations of the current model should be noted. First, the thermodynamic state
branch introduces a degree of flexibility that, if unconstrained, could in principle absorb
variance attributable to chemical abundances. The leakage test
(Section~\ref{subsec:leakage}) bounds this risk quantitatively.

Second, the imputation procedure used to fill missing ASPCAP labels introduces a
correlation structure into the training labels that may not reflect the true astrophysical
covariance. This could bias the thermodynamic state vector for stars with many missing
labels. A more principled treatment, such as training with an explicit latent variable
model for missing data, is a natural extension.

Third, unlike The Cannon 2, which delivers formal label uncertainties via the Jacobian
of the test-step likelihood, our current model does not yet support backward inference
(the recovery of labels from an observed spectrum). A planned backward model
(Section~\ref{sec:future}) will close this gap by using the emulator as a forward model
within an optimization loop.

The PCA manifold analysis demonstrates that the generated spectra are statistically
indistinguishable from real APOGEE spectra in the first two principal components. However,
the slightly higher cumulative explained variance for generated spectra (because they lack
observational noise) means that the comparison is not perfectly fair at high principal
component order. A more rigorous test, using a trained discriminative classifier to
distinguish real from generated spectra, is planned as part of the external validation
program described below.

%------------------------------------------------------------
\section{Future Work}
\label{sec:future}
%------------------------------------------------------------

We outline the following planned extensions and validation experiments.

\paragraph{External validation via classifier.}
A discriminative neural network will be trained to distinguish real APOGEE spectra from
emulator-generated ones. If the classifier's accuracy is near chance level (50\%), it
will confirm that generated and real spectra are statistically indistinguishable.

\paragraph{Backward inference pipeline.}
The emulator will be embedded in an optimization loop in which observed spectra serve
as targets. Stellar labels will be initialized from ASPCAP values and refined by
gradient descent on the forward model. Because the emulator runs in $\mathcal{O}$(ms),
this approach will be orders of magnitude faster than Markov Chain Monte Carlo methods
while retaining full uncertainty quantification if combined with an ensemble or variational
approach. A backward-model notebook is under active development.

\paragraph{Synthetic dataset generator.}
A public Python class will allow users to specify a parameter space and generate arbitrary
numbers of synthetic spectra for downstream applications such as training data-driven
classifiers or testing spectral analysis pipelines.

%------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
%------------------------------------------------------------

We have constructed a physics-constrained generative neural emulator for APOGEE DR17
stellar spectra. The model is trained on 149,966 real giant-star observations and achieves
a median reduced $\chi^2_\mathrm{red} = 1.328$ on a held-out test set of $\sim$9{,}500
stars, outperforming all four baselines (Table~\ref{tab:chi2}). By decomposing the
predicted optical depth into physically motivated branches and enforcing element-specific
weighted spectral masks supplemented by a stochastic Jacobian penalty, the model
guarantees that each chemical abundance label affects only those pixels associated with
its known atomic transitions. Sensitivity gradients and a quantitative information-leakage
test confirm this disentanglement at the pixel level.

The model outperforms The Payne, Expanded Payne, a baseline CNN, and RegPayne in
reduced $\chi^2$ while providing provable disentanglement that none of the unconstrained
baselines achieve. It represents a step toward spectral emulators that are simultaneously
accurate, physically transparent, and faithful to the observational properties of the
instruments that collect the data.

All code for this project is available, with documentation and examples, via the project
repository. Community contributions and discussion via GitHub are warmly welcomed.

%------------------------------------------------------------
\begin{authorcontributions}
A.S. conceived the model architecture, collected and preprocessed the data, performed
all training and validation experiments, produced all figures, and wrote the manuscript.
\end{authorcontributions}

\begin{acknowledgments}
%------------------------------------------------------------

The author thanks the supervisors and collaborators at Cepheid, IITH for guidance and
feedback. This project made use of data from the SDSS-IV APOGEE survey. Funding for
the Sloan Digital Sky Survey IV has been provided by the Alfred P.\ Sloan Foundation,
the U.S.\ Department of Energy Office of Science, and the Participating Institutions.
SDSS-IV acknowledges support and resources from the Center for High-Performance Computing
at the University of Utah. The SDSS web site is \texttt{www.sdss.org}.

This research made use of Astropy, a community-developed core Python package for
Astronomy \citep{Astropy2013}; NumPy \citep{vanderWalt2011}; SciPy \citep{Jones2001};
Matplotlib \citep{Hunter2007}; and scikit-learn \citep{Pedregosa2011}.
\end{acknowledgments}
%------------------------------------------------------------
% Use BibTeX with the v7 style file (aasjournalv7.bst)
% Place your .bib file references below, or use:
%   \bibliographystyle{aasjournalv7}
%   \bibliography{references}
%
% Manual bibliography shown here for portability:
\begin{thebibliography}{}

\bibitem[Abdurrouf et al.(2022)]{Abdurrouf2022}
Abdurrouf, et al.\ 2022, ApJS, 259, 35

\bibitem[Astropy Collaboration et al.(2013)]{Astropy2013}
Astropy Collaboration, Robitaille, T.\ P., Tollerud, E.\ J., et al.\ 2013, A\&A, 558, AA33

\bibitem[Bland-Hawthorn \& Gerhard(2016)]{BlandHawthorn2016}
Bland-Hawthorn, J., \& Gerhard, O.\ 2016, ARA\&A, 54, 529

\bibitem[Casey et al.(2016)]{Casey2016Cannon2}
Casey, A.\ R., Hogg, D.\ W., Ness, M., Rix, H.-W., Ho, A.\ Y.\ Q., \& Gilmore, G.\
2016, arXiv:1603.03040

\bibitem[Freeman \& Bland-Hawthorn(2002)]{Freeman2002}
Freeman, K., \& Bland-Hawthorn, J.\ 2002, ARA\&A, 40, 487

\bibitem[Garc\'ia P\'erez et al.(2016)]{GarciaperezASPCAP2016}
Garc\'ia P\'erez, A.\ E., Allende Prieto, C., Holtzman, J.\ A., et al.\ 2016,
AJ, 151, 144

\bibitem[Hunter(2007)]{Hunter2007}
Hunter, J.\ D.\ 2007, Comput.\ Sci.\ Eng., 9, 90

\bibitem[Jones et al.(2001)]{Jones2001}
Jones, E., Oliphant, E., Peterson, P., et al.\ 2001, SciPy: Open Source Scientific
Tools for Python, \url{http://www.scipy.org/}

\bibitem[Majewski et al.(2017)]{Majewski2017}
Majewski, S.\ R., Schiavon, R.\ P., Frinchaboy, P.\ M., et al.\ 2017, AJ, 154, 94

\bibitem[Ness et al.(2015)]{Ness2015Cannon}
Ness, M., Hogg, D.\ W., Rix, H.-W., Ho, A.\ Y.\ Q., \& Zasowski, G.\ 2015,
ApJ, 808, 16

\bibitem[Pedregosa et al.(2011)]{Pedregosa2011}
Pedregosa, F., Varoquaux, G., Gramfort, A., et al.\ 2011, JMLR, 12, 2825

\bibitem[Shetrone et al.(2015)]{Shetrone2015}
Shetrone, M., Bizyaev, D., Lawler, J.\ E., et al.\ 2015, ApJS, 221, 24

\bibitem[Smith et al.(2021)]{Smith2021DR17}
Smith, V.\ V., Cunha, K., Shetrone, M.\ D., et al.\ 2021, AJ, 161, 254

\bibitem[Ting et al.(2019)]{Ting2019Payne}
Ting, Y.-S., Conroy, C., Rix, H.-W., \& Cargile, P.\ 2019, ApJ, 879, 69

\bibitem[TransformerPayne(2025)]{TransformerPayne2025}
TransformerPayne Collaboration\ 2025, arXiv preprint [astro-ph.SR]

\bibitem[van der Walt et al.(2011)]{vanderWalt2011}
van der Walt, S., Colbert, S.\ C., \& Varoquaux, G.\ 2011, Comput.\ Sci.\ Eng., 13, 22

\end{thebibliography}

\end{document}